DE Analysis

DESeq2 analysis for Aedes aegypti RNAi knockdown RNA-seq
Runs the SAME analysis from two count sources so results are comparable:
   1. featureCounts  -> DESeqDataSetFromMatrix
   2. Salmon/tximport -> DESeqDataSetFromTximport (preserves length offsets)

 Models fit for each source:
   - one per tissue x sex group          (design auto-simplifies to ~ condition)
   - all whole tissues combined          (design ~ tissue_sex + condition)
   - cell line alone                     (design ~ condition)

```{r}
library(here)
library(DESeq2)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tibble)
library(purrr)
library(readr)
library(ashr)
library(tximport)
```

```{r}
# need to make txi.rds again
old_salmon_quant_dir <- "/projectnb/lau-bumc/emily/NF-Pipeline/mosquito_tissues_old_results/salmon_quant"

tx2gene_file <- "/projectnb/lau-bumc/emily/NF-Pipeline/bin/tx2gene.tsv"
tx2gene <- read_tsv(tx2gene_file, col_names = c("TXNAME", "GENEID"))

# quant.sf files are staged into the work dir (current dir)
files <- list.files(old_salmon_quant_dir, pattern = "quant.sf$", recursive = TRUE, full.names = TRUE)
names(files) <- sub("_salmon.*", "", basename(dirname(files)))
stopifnot(length(files) > 0, all(file.exists(files)))

txi <- tximport(files,
                type = "salmon",
                tx2gene = tx2gene,
                ignoreTxVersion = TRUE,
                countsFromAbundance = "no")

saveRDS(txi, here("mosq_old_txi.rds"))
```


```{r}
# NOTE: FeatureCounts excluded multi-mapped reads

fc_old <- "/projectnb/lau-bumc/emily/NF-Pipeline/mosquito_tissues_old_results/feature_counts/counts_matrix.txt"
#fc_rpt <- "/projectnb/lau-bumc/emily/NF-Pipeline/mosquito_tissues_rpt_results/feature_counts/counts_matrix.txt"
fc_cl <- "/projectnb/lau-bumc/emily/NF-Pipeline/siRNA_results/feature_counts/counts_matrix.txt"
fc_list <- c(fc_old, fc_cl)

slmn_old <- "/projectnb/lau-bumc/emily/mosq_old_txi.rds"
#slmn_rpt <- "/projectnb/lau-bumc/emily/NF-Pipeline/mosquito_tissues_rpt_results/salmon_counts/txi.rds"
slmn_cl <- "/projectnb/lau-bumc/emily/NF-Pipeline/siRNA_results/salmon_counts/txi.rds"
slmn_list <- c(slmn_old, slmn_cl)
```

```{r}
# output directory
outdir <- here("DE_results")

dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

```

```{r}
# test
sapply(fc_list, file.exists)
sapply(slmn_list, file.exists)
```

===========================================================================
Helpers
===========================================================================

Build the experimental design table from sample names -----------------
Used for BOTH count sources so the two analyses share an identical design.

```{r}
build_coldata <- function(samples) {
  cd <- data.frame(
    row.names = samples,

    condition = dplyr::case_when(
      grepl("siGFP|WT",  samples) ~ "WT",     # note: single string, alternation INSIDE
      grepl("AGO2|AG2",  samples) ~ "AGO2",
      grepl("AGO1",      samples) ~ "AGO1",
      grepl("R2D2",      samples) ~ "R2D2",
      grepl("DCR2",      samples) ~ "DCR2",
      grepl("DCR1",      samples) ~ "DCR1",
      TRUE ~ NA_character_
    ),

    sex = dplyr::case_when(
      grepl("Aag2", samples) ~ "none",        # cell line has no sex
      grepl("_M_",  samples) ~ "M",
      grepl("_F_",  samples) ~ "F",
      TRUE ~ NA_character_
    ),

    tissue = dplyr::case_when(
      grepl("Aag2",   samples) ~ "cell_line", # keyed on Aag2, NOT on condition strings
      grepl("carc",   samples) ~ "carc",
      grepl("head",   samples) ~ "head",
      grepl("testes", samples) ~ "testes",
      grepl("ovary",  samples) ~ "ovary",
      TRUE ~ NA_character_
    ),

    # renamed from `library` to avoid shadowing base::library()
    lib_type = ifelse(grepl("ttlRNA", samples), "ttlRNA", "LRNA"),

    stringsAsFactors = FALSE
  )

  # combined tissue+sex factor: avoids rank-deficiency from the partial
  # confound (ovary is F-only, testes is M-only) in the combined model
  cd$tissue_sex <- paste(cd$tissue, cd$sex, sep = "_")

  cd$condition  <- relevel(factor(cd$condition), ref = "WT")
  cd$tissue     <- factor(cd$tissue)
  cd$sex        <- factor(cd$sex)
  cd$lib_type   <- factor(cd$lib_type)
  cd$tissue_sex <- factor(cd$tissue_sex)

  # loud failure if any sample didn't parse
  if (any(is.na(cd$condition)) || any(is.na(cd$tissue))) {
    warning("Some samples failed to parse:\n",
            paste(rownames(cd)[is.na(cd$condition) | is.na(cd$tissue)], collapse = "\n"))
  }
  cd
}
```

Subset a tximport object -----------------------------------------------
THIS IS THE FIX for the Salmon bug: txi is a LIST of matrices, so it
cannot be subset with txi[, samples]. Each component must be subset,
and countsFromAbundance must be carried through or DESeq2 errors.

```{r}
subset_txi <- function(txi, samples) {
  list(
    abundance           = txi$abundance[, samples, drop = FALSE],
    counts              = txi$counts[,    samples, drop = FALSE],
    length              = txi$length[,    samples, drop = FALSE],
    countsFromAbundance = txi$countsFromAbundance
  )
}
```
Drop design terms that have <2 levels in this subset --------------------
Lets one master formula (~ tissue_sex + condition) work for every group:
within a single tissue x sex group, tissue_sex collapses to one level and
is silently dropped, leaving ~ condition.
```{r}
simplify_design <- function(coldata_sub, design_vars) {
  keep <- design_vars[
    vapply(design_vars,
           function(v) nlevels(droplevels(factor(coldata_sub[[v]]))) >= 2,
           logical(1))
  ]
  if (!("condition" %in% keep)) return(NULL)         # nothing to test
  keep <- c(setdiff(keep, "condition"), "condition") # condition LAST, by convention
  as.formula(paste("~", paste(keep, collapse = " + ")))
}
```

Core DE runner ----------------------------------------------------------
count_src is either a plain counts matrix (featureCounts) or a txi list
(Salmon). Dispatches to the correct DESeq2 constructor.

```{r}
run_de <- function(sample_ids, coldata, count_src, design_vars, label) {

  coldata_sub <- droplevels(coldata[sample_ids, , drop = FALSE])

  # --- pre-flight: enough replication to fit a model? ---
  rep_table <- table(coldata_sub$condition)

  if (!("WT" %in% names(rep_table)) ||
      rep_table[["WT"]] < 2 ||
      sum(rep_table >= 2) < 2) {
    message("  [skip] ", label, ": insufficient replication (need WT>=2 and >=2 conditions with >=2 reps)")
    return(NULL)
  }

  # drop under-replicated conditions (n=1) so they don't break dispersion estimation
  ok_conds <- names(rep_table)[rep_table >= 2]
  dropped  <- setdiff(names(rep_table)[rep_table > 0], ok_conds)
  if (length(dropped)) {
    message("  [note] ", label, ": dropping n<2 conditions: ", paste(dropped, collapse = ", "))
  }

  coldata_sub <- droplevels(
    coldata_sub[coldata_sub$condition %in% ok_conds, , drop = FALSE]
  )
  coldata_sub$condition <- relevel(factor(coldata_sub$condition), ref = "WT")
  keep_ids <- rownames(coldata_sub)

  # --- design ---
  design_formula <- simplify_design(coldata_sub, design_vars)
  if (is.null(design_formula)) {
    message("  [skip] ", label, ": no usable design")
    return(NULL)
  }
  message("  [run ] ", label, "  n=", length(keep_ids),
          "  design: ", deparse(design_formula))

  # --- build the DESeqDataSet from whichever source ---
  dds <- tryCatch({
    if (is.list(count_src) && !is.null(count_src$length)) {
      # Salmon / tximport: subset the txi, keep the length offsets
      txi_sub <- subset_txi(count_src, keep_ids)
      DESeqDataSetFromTximport(txi_sub,
                               colData = coldata_sub,
                               design  = design_formula)
    } else {
      # featureCounts: plain integer matrix
      counts_sub <- count_src[, keep_ids, drop = FALSE]
      DESeqDataSetFromMatrix(countData = counts_sub,
                             colData   = coldata_sub,
                             design    = design_formula)
    }
  }, error = function(e) {
    message("  [fail] ", label, " (construct): ", conditionMessage(e))
    NULL
  })
  if (is.null(dds)) return(NULL)

  dds <- tryCatch(DESeq(dds), error = function(e) {
    message("  [fail] ", label, " (DESeq): ", conditionMessage(e))
    NULL
  })
  if (is.null(dds)) return(NULL)

  # --- shrunken results for every knockdown present, vs WT ---
  conds <- setdiff(levels(coldata_sub$condition), "WT")

  contrasts <- set_names(
    map(conds, function(cond) {
      lfcShrink(dds,
                contrast = c("condition", cond, "WT"),
                type     = "ashr") |>       # ashr required: apeglm can't use contrast=
        as.data.frame() |>
        rownames_to_column("gene_id") |>
        arrange(padj)
    }),
    paste0(conds, "_vs_WT")
  )

  c(list(dds = dds), contrasts)
}
```

```{r}
## Export dds + one CSV per contrast --------------------------------------
export_results <- function(res_list, group_name, source_tag) {
  if (is.null(res_list)) return(invisible(NULL))

  saveRDS(res_list$dds,
          file.path(outdir, paste0(source_tag, "_", group_name, "_dds.rds")))

  contrast_names <- setdiff(names(res_list), "dds")
  walk(contrast_names, function(cn) {
    write_csv(res_list[[cn]],
              file.path(outdir, paste0(source_tag, "_", group_name, "_", cn, ".csv")))
  })
  invisible(NULL)
}
```

```{r}
## Significant genes: filtered table + one-row summary --------------------
## Thresholds are set once here so every contrast is called the same way.
PADJ_CUT <- 0.05
LFC_CUT  <- 1

export_sig_genes <- function(res_df, group_name, contrast, source_tag,
                             padj_cut = PADJ_CUT, lfc_cut = LFC_CUT) {

  sig <- res_df |>
    filter(!is.na(padj), padj < padj_cut, abs(log2FoldChange) > lfc_cut) |>
    mutate(direction = ifelse(log2FoldChange > 0, "up", "down")) |>
    arrange(padj)

  write_csv(sig,
            file.path(outdir,
                      paste0(source_tag, "_", group_name, "_", contrast, "_SIGNIFICANT.csv")))

  # one-row summary for the master table
  tibble(
    source    = source_tag,
    group     = group_name,
    contrast  = contrast,
    n_tested  = sum(!is.na(res_df$padj)),          # genes that survived independent filtering
    n_sig     = nrow(sig),
    n_up      = sum(sig$direction == "up"),
    n_down    = sum(sig$direction == "down"),
    padj_cut  = padj_cut,
    lfc_cut   = lfc_cut,
    top_gene  = if (nrow(sig)) sig$gene_id[1] else NA_character_,
    top_lfc   = if (nrow(sig)) round(sig$log2FoldChange[1], 3) else NA_real_
  )
}
```

```{r}
## PCA plot ---------------------------------------------------------------
plot_pca <- function(dds, label, source_tag, color_by = "condition", shape_by = NULL) {
  vsd <- vst(dds, blind = TRUE)

  intgroup <- unique(c(color_by, shape_by, "tissue_sex"))
  intgroup <- intgroup[intgroup %in% colnames(colData(dds))]

  pca_data    <- plotPCA(vsd, intgroup = intgroup, returnData = TRUE)
  percent_var <- round(100 * attr(pca_data, "percentVar"))

  p <- ggplot(pca_data,
              aes(x = PC1, y = PC2,
                  color = .data[[color_by]],
                  label = name)) +
    geom_point(size = 3) +
    geom_text_repel(size = 2.5, show.legend = FALSE, max.overlaps = 30) +
    xlab(paste0("PC1: ", percent_var[1], "% variance")) +
    ylab(paste0("PC2: ", percent_var[2], "% variance")) +
    ggtitle(paste0(source_tag, " — ", label)) +
    theme_bw()

  if (!is.null(shape_by) && shape_by %in% colnames(pca_data)) {
    p <- p + aes(shape = .data[[shape_by]])
  }

  ggsave(file.path(outdir, paste0(source_tag, "_", label, "_PCA.png")),
         p, width = 8, height = 6, dpi = 150)
  invisible(p)
}
```

```{r}
## Volcano plot -----------------------------------------------------------
plot_volcano <- function(res_df, label, contrast, source_tag,
                         padj_cut = PADJ_CUT, lfc_cut = LFC_CUT) {
  df <- res_df |>
    filter(!is.na(padj)) |>
    mutate(sig = case_when(
      padj < padj_cut & log2FoldChange >  lfc_cut ~ "up",
      padj < padj_cut & log2FoldChange < -lfc_cut ~ "down",
      TRUE ~ "ns"
    ))

  top <- df |> filter(sig != "ns") |> slice_head(n = 20)

  p <- ggplot(df, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
    geom_point(alpha = 0.6, size = 1.2) +
    geom_vline(xintercept = c(-lfc_cut, lfc_cut), linetype = "dashed", color = "grey60") +
    geom_hline(yintercept = -log10(padj_cut),     linetype = "dashed", color = "grey60") +
    geom_text_repel(data = top, aes(label = gene_id),
                    size = 2.5, max.overlaps = 30, show.legend = FALSE) +
    scale_color_manual(values = c(up = "firebrick", down = "steelblue", ns = "grey70")) +
    labs(title = paste0(source_tag, " — ", label, " — ", contrast),
         x = "log2 fold change (shrunken)", y = "-log10 adjusted p") +
    theme_bw()

  ggsave(file.path(outdir, paste0(source_tag, "_", label, "_", contrast, "_volcano.png")),
         p, width = 7, height = 6, dpi = 150)
  invisible(p)
}

```

```{r}
## Run every model for one count source ----------------------------------
run_all_models <- function(count_src, coldata, source_tag) {

  message("\n=== ", source_tag, " ===")

  # ---- 1. tissue x sex groups (derived from the data, not hardcoded) ----
  ts_groups <- coldata |>
    rownames_to_column("sample") |>
    filter(!is.na(tissue), !is.na(sex)) |>
    mutate(grp = paste(tissue, sex, sep = "_")) |>
    split(~ grp) |>
    map(~ .x$sample)

  # ---- 2. combined models: whole tissues pooled, cell line separate ----
  combined_groups <- list(
    all_tissues = rownames(coldata)[coldata$tissue != "cell_line"],
    cell_line   = rownames(coldata)[coldata$tissue == "cell_line"]
  )
  combined_groups <- combined_groups[lengths(combined_groups) > 0]

  all_groups <- c(ts_groups, combined_groups)

  # master design; simplify_design() prunes single-level terms per group, so
  #   - within one tissue x sex group -> ~ condition
  #   - all_tissues                   -> ~ tissue_sex + condition
  #   - cell_line                     -> ~ condition
  design_vars <- c("tissue_sex", "condition")

  results <- imap(all_groups, function(sample_ids, group_name) {
    run_de(sample_ids, coldata, count_src, design_vars, group_name)
  })

  results <- compact(results)   # drop the NULLs from skipped groups

  # ---- export: dds, full results, significant genes, plots ----
  summaries <- imap(results, function(res_list, group_name) {

    export_results(res_list, group_name, source_tag)

    plot_pca(res_list$dds, group_name, source_tag,
             color_by = "condition",
             shape_by = if (group_name == "all_tissues") "tissue_sex" else NULL)

    contrast_names <- setdiff(names(res_list), "dds")

    map_dfr(contrast_names, function(cn) {
      plot_volcano(res_list[[cn]], group_name, cn, source_tag)
      export_sig_genes(res_list[[cn]], group_name, cn, source_tag)   # returns 1-row summary
    })
  })

  summary_df <- bind_rows(summaries)

  list(results = results, summary = summary_df)
}

```

```{r}
# ===========================================================================
# 1. featureCounts
# ===========================================================================
fc_out_list <- list()
fc_results_list <- list()  # collect everything here — usually easier than loose global objects


for (i in seq_along(fc_list)) {

  featureCounts_output <- fc_list[[i]]     # the current file path

  fc <- read.delim(featureCounts_output,
                    comment.char = "#",
                    check.names  = FALSE)

  fc_counts <- as.matrix(fc[, 7:ncol(fc)])
  rownames(fc_counts) <- fc$Geneid
  colnames(fc_counts) <- gsub("\\.dedup\\.bam$", "", basename(colnames(fc_counts)))

  fc_coldata <- build_coldata(colnames(fc_counts))
  stopifnot(identical(rownames(fc_coldata), colnames(fc_counts)))
  
  # give this iteration's result a unique name
  out_name <- paste0("fc_out_", i)
  result_name <- paste0("fc_results_", i)
  
  fc_out     <- run_all_models(fc_counts, fc_coldata, "featureCounts")
  fc_results <- fc_out$results

  # also store in a list, keyed by name
  fc_out_list[[out_name]] <- fc_out
  fc_results_list[[result_name]] <- fc_results
}
```

```{r}
# tests
```

```{r}
# ===========================================================================
# 2. Salmon / tximport
# ===========================================================================
salmon_out_list <- list()
salmon_results_list <- list()

for (i in seq_along(slmn_list)) {
  
  txi <- slmn_list[[i]]
  txi <- readRDS(txi)
  
  salmon_coldata <- build_coldata(colnames(txi$counts))
  stopifnot(identical(rownames(salmon_coldata), colnames(txi$counts)))
  
  salmon_out     <- run_all_models(txi, salmon_coldata, "salmon")
  salmon_results <- salmon_out$results
  
  out_name <- paste0("salmon_out_", i)
  result_name <- paste0("salmon_results_", i)
  
  salmon_out_list[[out_name]] <- salmon_out
  salmon_results_list[[result_name]] <- salmon_results
  
}
```

```{r}
# tests
fc_samples     <- colnames(fc_results_list[[1]]$all_tissues$dds)
salmon_samples <- colnames(salmon_results_list[[1]]$all_tissues$dds)

setdiff(fc_samples, salmon_samples)   # sample(s) present in fc but dropped from salmon
setdiff(salmon_samples, fc_samples)   # should be empty

```


```{r}
# ===========================================================================
# 3. Master summary table across both sources
# ===========================================================================

for (i in seq_along(fc_results_list)) {
  # get fc and salmon results from results lists
  fc_out <- fc_out_list[[i]]
  salmon_out <- salmon_out_list[[i]]
  
  de_summary <- bind_rows(fc_out$summary, salmon_out$summary) |>
    arrange(source, group, contrast)
  
  filename <- paste0("DE_summary_all_", i )
  write_csv(de_summary, file.path(outdir, filename))
  
  message("\n--- Significant genes (padj < ", PADJ_CUT,
          ", |log2FC| > ", LFC_CUT, ") ---")
  print(as.data.frame(de_summary[, c("source", "group", "contrast",
                                     "n_tested", "n_sig", "n_up", "n_down")]),
        row.names = FALSE)
}

```

```{r}
# ===========================================================================
# 4. Sample-name reconciliation check (needed for the two-method comparison)
# ===========================================================================

only_fc  <- setdiff(colnames(fc_counts),  colnames(txi$counts))
only_sal <- setdiff(colnames(txi$counts), colnames(fc_counts))

if (length(only_fc) || length(only_sal)) {
  message("\n[WARN] Sample names differ between featureCounts and Salmon.")
  if (length(only_fc))  message("  featureCounts only: ", paste(only_fc,  collapse = ", "))
  if (length(only_sal)) message("  Salmon only:        ", paste(only_sal, collapse = ", "))
  message("  Normalise these before comparing DEG lists across methods.")
}

message("\nDone. Outputs written to: ", outdir)
```
Look up significant genes
```{r}
# mosquito tissue genes
salmon_results_list[[1]]
fc_results_list[[1]]

# cell line genes
salmon_results_list[[2]]
fc_results_list[[2]]
```
```{r}
# mosquito male carcass tissues
#salmon_results_list[[1]]$carc_M$AGO2_vs_WT
#fc_results_list[[1]]$carc_M$AGO2_vs_WT

sig1 <- salmon_results_list[[1]]$carc_M$AGO2_vs_WT |>
  filter(!is.na(padj), padj < PADJ_CUT, abs(log2FoldChange) > LFC_CUT) |>
  mutate(direction = ifelse(log2FoldChange > 0, "up", "down")) |>
  arrange(padj)
sig1.2 <- fc_results_list[[1]]$carc_M$AGO2_vs_WT |>
  filter(!is.na(padj), padj < PADJ_CUT, abs(log2FoldChange) > LFC_CUT) |>
  mutate(direction = ifelse(log2FoldChange > 0, "up", "down")) |>
  arrange(padj)
intersect(sig1$gene_id, sig1.2$gene_id)
length(sig1$gene_id)
length(sig1.2$gene_id)
# mosquito female carcass tissues (iffy)
#salmon_results_list[[1]]$carc_F$AGO2_vs_WT
#fc_results_list[[1]]$carc_F$AGO2_vs_WT

sig2 <- salmon_results_list[[1]]$carc_F$AGO2_vs_WT |>
  filter(!is.na(padj), padj < PADJ_CUT, abs(log2FoldChange) > LFC_CUT) |>
  mutate(direction = ifelse(log2FoldChange > 0, "up", "down")) |>
  arrange(padj)

# cell line
#salmon_results_list[[2]]$cell_line$AGO2_vs_WT
#fc_results_list[[2]]$cell_line$AGO2_vs_WT

sig3 <- salmon_results_list[[2]]$cell_line$AGO2_vs_WT |>
  filter(!is.na(padj), padj < PADJ_CUT, abs(log2FoldChange) > LFC_CUT) |>
  mutate(direction = ifelse(log2FoldChange > 0, "up", "down")) |>
  arrange(padj)

sig_genes_M_F <- intersect(sig1$gene_id, sig2$gene_id)
sig_genes_M_cl <- intersect(sig1$gene_id, sig3$gene_id)
sig_genes_F_cl <- intersect(sig2$gene_id, sig3$gene_id)

sig_all_tiss <- salmon_results_list[[1]]$all_tissues$AGO2_vs_WT |>
  filter(!is.na(padj), padj < PADJ_CUT, abs(log2FoldChange) > LFC_CUT) |>
  mutate(direction = ifelse(log2FoldChange > 0, "up", "down")) |>
  arrange(padj)

intersect(sig3$gene_id, sig_all_tiss$gene_id)
```

```{r}
# feature counts genes
# all_tiss_ago2

fc_all_tiss_ago2 <- read_csv("/projectnb/lau-bumc/emily/DE_results/results_to_use/featureCounts_genes/featureCounts_all_tissues_AGO2_vs_WT_SIGNIFICANT.csv")

fc_cell_ago2 <- read_csv("/projectnb/lau-bumc/emily/DE_results/results_to_use/featureCounts_genes/featureCounts_cell_line_AGO2_vs_WT_SIGNIFICANT.csv")

# salmon genes

slmn_all_tiss_ago2 <- read_csv("/projectnb/lau-bumc/emily/DE_results/results_to_use/Salmon_genes/salmon_all_tissues_AGO2_vs_WT_SIGNIFICANT.csv")

slmn_cell_ago2 <- read_csv("/projectnb/lau-bumc/emily/DE_results/results_to_use/Salmon_genes/salmon_cell_line_AGO2_vs_WT_SIGNIFICANT.csv")
```
```{r}
# create correlation plots


```



